Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-08-25 manuscript/usses_water_ms.tex
adds bib file

\title{Consistent, and weak, ecosystem functional response across precipitation
extremes in a sagebrush steppe}

@@

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to availabile soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also no evidence that treatment effects grew
over time. Plant community composition was remarkably stable over the
course of the experiment and did not differ among treatments.

\textbf{Discussion.} Despite some evidence that ecosystem functional
response saturated under extreme wet conditions, the response of ANPP to
soil moisture was consistintly weak and community composition was
remarkably stable. The similarity, or difference, of ecosystem
functional response across treatments was not due to compensatory shifts
at the plant community level, but instead reflects the weak responses of
individual species. Such responses may be due to bet-hedging strategies
where species respond weakly to even extreme levels of soil moisture to
ensure low variance of long-term success.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions. Rainy days were
defined as any day in which precipitation was recorded and average
temperature was above 3\(^{\circ}\)C. The day immediately following
rainfall was also classified as rainy. We fit the model using the
`lme4::lmer()' function \citep{Bates2015} in R \citep{R2016}, with
random effects for plot group and date. We weighted observations by the
number of unique sensors or spot measurements that were taken in each
plot on that day. We then used the model to predict the average daily
soil moisture in the treated plots based on the average daily ambient
soil moisture. We could only predict soil moisture in the treated plots
on days for which we took at least one ambient soil moisture
measurement.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR bands for NDVI. To
convert the greenness index to ANPP we regressed the greenness index
against the dry biomass weight from the ten calibration plots. We fit
regressions to MODIS-based index and AVHRR-based index for each year and
retained the regression with the better fit. Using the best regression
equation for each year, we predicted ANPP (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. From those maps, we have data on the
density of all annuals and perennials forbs, basal cover of perennial
grasses, and canopy cover of shrubs. We made a large plot-treatment-year
by species matrix, where columns were filled with either basal cover or
density, depending on the measurement made for the particular species.
So we could analyze the different types of data together, we
standardized the values in each column. This puts all abundance values
on the same scale, meaning that common and rare species are weighted
equally. Nonetheless, if we assume that rare species will respond more
than common ones, then this approach is anti-conservative for detecting
community change. This means that our approach is biased toward
detecting compositional changes.

@@

\begin{description}
\item [H1] The coefficient for drought$\times$VWC is positive and different from zero.
\item [H2] The coefficient for irrigation$\times$VWC is negative and different from zero.
\end{description}

To account for the fact that observations within plots and years are not
independent, we include two random effects. Specifically, we include
plot-specific offsets (\(\boldsymbol{\gamma}\)) for the intercept and
slope terms and year-specific intercept offsets (\(\eta_t\)). The
covariate vector \(\textbf{z}_i\) for each observation \emph{i} has two
elements: a 1 for the intercept and the scaled value of VWC for that
plot and year. The plot-specific coefficients are modeled
hierarchically, where plot level coefficients are drawn from a
multivariate normal distribution with mean 0 and a variance-covariance
strucutre that allows the intercept and slope terms to be correlated:

@@

We fit the model using a Bayesian approach, obtaining posterior
estimates of all unknowns via the No-U-Turn Hamiltonian Monte Carlo
sampler in Stan \citep{stan2016}. We used the R package `rstan'
\citep{rstan2016} to link R \citep{R2016} to Stan. We obtained samples
from the posterior distribution for all model parameters from four
parallel MCMC chains run for 10,000 iterations, saving every
\(10^{\text{th}}\) sample. Trace plots of all parameters were visually
inspected to ensure well-mixed chains and convergence. We also made sure
all scale reduction factors (\(\hat{R}\)) were less than 1.1.

To see if community composition differed among treatments through time,
we used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis
distances. For each year of the experiment, we first calculated
Bray-Curtis distances among all plots, and then extracted those
distances for use in the NMDS. Because we standardized species'
abundances, some values were negative, which is not allowed for
calculating Bray-Curtis distances. We simply added `2' to each abundance
value to ensure all values were greater than zero. We plotted the first
two axes of NMDS scores to see if community composition overlapped, or
not, among treatments in each year. We used the `metaMDS()' function in
the R package `vegan' \citep{Oksanen2016} to calculate Bray-Curtis
distances and then to run the NMDS analysis. We used the
`vegan::adonis()' function \citep{Oksanen2016} to perform permutational
multivariate analysis of variance to test whether treatment plots formed
distinct groupings. To test whether treatment plots were equally
dispersed, or not, we used the `vegan::betadisper' function
\citep{Oksanen2016}.

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3). The effect of soil moisture for each treatment is
associated with high uncertainty, however, with credible intervals that
overlap zero (Table 1). Although the parameter estimates for the effect
of soil moisture overlap zero, the posterior distributions of the slopes
all shrank and shifted to more positive values relative to the prior
distributions (Fig. A2-2). Ecosystem functional response was similar
among treatments (Table 1; Fig. 3B), but there is evidence that the
slope for the irrigation treatment is less than the slope for the
control treatment. This evidence comes from interpreting the posterior
distribution of the slope offset for the irrigation treatment, from
which we calculate a 99\% one-tailed probability that the estimate is
less than zero (Fig. 3A, right panel). There was no evidence that the
treatment effects became more important over time because there was no
directional trend in the random year effects (Fig. A2-2).

Community composition was similar among treatments. In no year did
community composition among treatments not overlap, and they were
equally dispersed in all years (Table 2; Fig. 5). Likewise, community
composition was remarkably stable over time, with no evidence of
divergence among treatments (Table 2; Fig. 5).

@@

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Under chronic
drought, neither ecosystem functional response nor community composition
changed (Fig. 1, top left). But under irrigation, ecosystem functional
response was different from the control treatment while community
composition remained unchanged (Fig. 1, top right). Thus, we found some
evidence for a saturating response at high water availability (Fig. 3A,
right panel). However, the response of ANPP to soil moisture was
consitently weak, regardless of treatment (Table 1).

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 5) between drought and control treatments is
surprising because grasslands generally, and sagebrush steppe
specifically, are considered water-limited systems. Indeed, we expected
ecosystem functional response, community composition, or both to change
under the drought treatment, landing us in any box of Fig. 1
\emph{except} the top left. So, why did our drought treatment fail to
induce ecosystem or community responses? We can think of three reasons;
two are limitations of our study, and one is the life history traits of
the species in our focal communities. We first discuss the potential
limitations of our study, and then discuss the biological explanation.

First, it could be that our manipulations were not large enough to
induce a response. That is, maybe a 50\% decrease/increase in any given
year is not abnormal given our site's historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our manipulations \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation necessary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site. The 1\% quantile of
precipitation at our site is 110 mm, a 47\% reduction from the mean, and
the 99\% quantile is 414 mm, a 77\% increase from mean growing season
precipitation (Appendix 3). Thus, our drought treatment represented
extreme precipitation amounts, especially in years where ambient
precipitation was below average (Fig. 2A). The irrigation treatment may
have been too small, yet that is the treatment where we observed an
effect (Fig. 3).

Second, ANPP at our site may be influenced by factors beyond the window
of soil moisture we included in our statistical model. For example,
temperature can impact ANPP directly \citep{Epstein1997} and by
exacerbating the effects of soil moisture \citep{DeBoeck2011}.
Measurements of soil moisture likely contain a signal of temperature,
through its impact on evaporation and infiltration, but the measurements
will not capture the direct effect of temperature on metabolic and
physiological processes. Likewise, we did not redistribute snow across
our treatments in the winter, and snow melt may spur early spring
growth. These statistical issues of missing potentially important
covariates could explain the weak and uncertain relationship we observed
between soil moisture and ANPP.

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistent but weak and uncertain effect of
soil moisture on ANPP (Fig. 3). Species that live in variable
environments, such as cold deserts, must have strategies to ensure
long-term success as conditions vary. One strategy is bet hedging, where
species forego short-term gains to reduce the variance of long-term
success \citep{Seger1987}. In other words, species do the same thing
every year, with only minimal response to environmental conditions. The
dry and variable environment of the sagebrush steppe has likely selected
for bet hedging species that can maintain function at low water
availability and have weak responses to high water availability. In so
doing, the dominant species in our ecosystem avoid ``boom and bust''
cycles, which corresponds to the weak effect of soil moisture on ANPP
(i.e., the credible intervals for \(\beta_1\) overlapping zero), even at
precipitation extremes.

Another strategy to deal with variable environmental conditions is
avoidance, which would result in a consistent ecosystem functional
response at low soil moisture. The perennial plants in this cold desert
ecosystem are tolerant to drought conditions \citep[A.R. Kleinhesselink,
unpublished data]{Bazzaz1979, Franks2011}. For example, many of the
perennial grasses in our focal ecosystem avoid drought stress by growing
early in the growing season \citep[A.R. Kleinhesselink, personal
observation]{Blaisdell1958}. Likewise, the dominant shrub in our focal
ecosystem, \emph{Artemisia tripartita}, has access to water deep in the
soil profile thanks to a deep root system \citep{Kulmatiski2017a}.

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because the
species in our focal plant community are tolerant of drought conditions
and bet hedgers in wet conditions, maintaining relatively consistent
responses to interannual variation in precipitation to avoid booms and
busts. Longer time series of chronic precipitation alteration may reveal
plant community shifts that we did not observe
\citep[e.g.,][]{Wilcox2016}, in which case species that do not bet hedge
may gain prominence and dominate the ecosystem functional response. Our
results suggest compositional shifts would have the largest impact at
high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.

>>>>

\title{Ecosystem functional response across precipitation extremes in a
sagebrush steppe}

@@

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to availabile soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large variation in ANPP within-years.
Plant community composition was remarkably stable over the course of the
experiment and did not differ among treatments.

\textbf{Discussion.} Despite some evidence that ecosystem functional
response was less positive under extreme wet conditions, the response of
ANPP to soil moisture was consistently weak and community composition
was remarkably stable. Ecosystem functional responses across treatments
was not related to compensatory shifts at the plant community level, but
instead reflects the weak and variable responses of individual species.
Such responses may be due to bet hedging strategies where species
respond weakly to even extreme levels of soil moisture to ensure low
variance of long-term success.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions. Rainy days were
defined as any day in which precipitation was recorded and average
temperature was above 3\(^{\circ}\)C. The day immediately following
rainfall was also classified as rainy. We fit the model using the
\texttt{lme4::lmer()} function \citep{Bates2015} in R \citep{R2016},
with random effects for plot group and date. We weighted observations by
the number of unique sensors or spot measurements that were taken in
each plot on that day. We then used the model to predict the average
daily soil moisture in the treated plots based on the average daily
ambient soil moisture. We could only predict soil moisture in the
treated plots on days for which we took at least one ambient soil
moisture measurement.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR bands for NDVI. We
regressed the greenness index against the dry biomass weight from the
ten calibration plots to convert the greenness index to ANPP. We fit
regressions to a MODIS-based index and an AVHRR-based index for each
year and retained the regression with the better fit based on \(R^2\)
values. We then predicted ANPP using the best regression equation for
each year (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. We extracted the mapped information to
obtain data on the density of all annuals and perennials forbs, the
basal cover of perennial grasses, and the canopy cover of shrubs. We
made a large plot-treatment-year by species matrix, where columns were
filled with either basal cover or density, depending on the measurement
made for the particular species. We standardized the values in each
column so we could analyze the different types of data together. This
puts all abundance values on the same scale, meaning that common and
rare species are weighted equally. This approach is anti-conservative
for detecting community change if we assume that rare species will
respond more than common ones, which biases our approach toward
detecting compositional changes.

@@

\begin{description}
\item [H1.] The coefficient for drought$\times$VWC is positive and different from zero.
\item [H2.] The coefficient for irrigation$\times$VWC is negative and different from zero.
\end{description}

We include two random effects to account for the fact that observations
within plots and years are not independent. Specifically, we include
plot-specific offsets (\(\boldsymbol{\gamma}\)) for the intercept and
slope terms and year-specific intercept offsets (\(\eta_t\)). The
covariate vector \(\textbf{z}_i\) for each observation \emph{i} has two
elements: a 1 for the intercept and the scaled value of VWC for that
plot and year. The plot-specific coefficients are modeled
hierarchically, where plot level coefficients are drawn from a
multivariate normal distribution with mean 0 and a variance-covariance
structure that allows the intercept and slope terms to be correlated:

@@

We fit the model using a Bayesian approach and obtained posterior
estimates of all unknowns via the No-U-Turn Hamiltonian Monte Carlo
sampler in Stan \citep{stan2016}. We used the R package `rstan'
\citep{rstan2016} to link R \citep{R2016} to Stan. We obtained samples
from the posterior distribution for all model parameters from four
parallel MCMC chains run for 10,000 iterations, saving every
\(10^{\text{th}}\) sample. Trace plots of all parameters were visually
inspected to ensure well-mixed chains and convergence. We also made sure
all scale reduction factors (\(\hat{R}\) values) were less than 1.1
\citep{Gelman2009}.

We used non-dimensional multivariate scaling (NMDS) based on Bray-Curtis
distances to see if community composition differed among treatments
through time. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3). The effect of soil moisture for each treatment is
associated with high uncertainty, however, with 95\% Bayesian credible
intervals that overlap zero (Table 1). Although the parameter estimates
for the effect of soil moisture overlap zero, the posterior
distributions of the slopes all shrank and shifted to more positive
values relative to the prior distributions (Fig. A2-2), which indicates
the data did influence parameter estimates beyond the information from
the uninformative priors. Ecosystem functional response was similar
among treatments (Table 1; Fig. 3B), but there is evidence that the
slope for the irrigation treatment is less than the slope for the
control treatment. This evidence comes from interpreting the posterior
distribution of the slope offset for the irrigation treatment, from
which we calculate a 99\% one-tailed probability that the estimate is
less than zero (Fig. 3A, right panel). There was no evidence that the
treatment effects became more important over time because there was no
directional trend in the random year effects (Fig. A2-3).

Community composition was similar among treatments. In no year did
community composition among treatments not overlap, and they were
equally dispersed in all years (Table 2; Fig. 5). Community composition
was also remarkably stable over time, with no evidence of divergence
among treatments (Table 2; Fig. 5).

@@

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Neither ecosystem
functional response nor community composition changed under chronic
drought (Fig. 3A, Fig. 4), consistent with the top left prediction in
Fig. 1. Ecosystem functional response under chronic irrigation was
different from the control treatment, but community composition remained
unchanged (Fig. 3A, Fig. 4), consistent with the top right prediction in
Fig. 1. The altered ecosystem functional response under irrigation
matched our expectations becuase we found evidence for a saturating
response for the irrigation treatment (Fig. 3A, right panel). However,
the response of ANPP to soil moisture in all treatments was consistently
weak (Table 1, Fig. A2-2) and within-year variation of ANPP was large
(Fig. 3B)

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 5) between drought and control treatments is
surprising because grasslands generally, and sagebrush steppe
specifically, are considered water-limited systems. Indeed, we expected
ecosystem functional response, community composition, or both to change
under the drought treatment, landing us in any box of Fig. 1
\emph{except} the top left. Why did our drought treatment fail to induce
ecosystem or community responses? We can think of three reasons; two are
limitations of our study, and one is the life history traits of the
species in our focal communities. We first discuss the potential
limitations of our study, and then discuss the biological explanation.

First, it could be that our drought manipulation was not large enough to
induce a response. That is, a 50\% decrease in any given year may not be
abnormal given our site's historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our drought treatment \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation necessary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site (Fig. A3-3). The 1\%
quantile of precipitation at our site is 110 mm, a 47\% reduction from
the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean
growing season precipitation (Appendix 3). Thus, our drought treatment
represented extreme precipitation amounts, especially in years where
ambient precipitation was below average (Fig. 2A). The irrigation
treatment may have been too small, yet that is the treatment where we
did observe an effect (Fig. 3).

Second, ANPP at our site may be influenced by factors beyond the window
of soil moisture (cumulative March-June) we included in our statistical
model. For example, temperature can impact ANPP directly
\citep{Epstein1997} and by exacerbating the effects of soil moisture
\citep{DeBoeck2011}. Measurements of soil moisture likely contain a
signal of temperature, through its impact on evaporation and
infiltration, but the measurements will not capture the direct effect of
temperature on metabolic and physiological processes. We also did not
redistribute snow across our treatments in the winter, and snow melt may
spur early spring growth. These statistical issues of missing
potentially important covariates could explain the within-year spread of
ANPP (Fig 2C, Fig. 3B) and the resulting uncertain relationship we
observed between soil moisture and ANPP across all treatments (Table 1,
Fig. A2-2).

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Fig. 3). Species that live in variable
environments, such as cold deserts, must have strategies to ensure
long-term success as conditions vary. One strategy is bet hedging, where
species forego short-term gains to reduce the variance of long-term
success \citep{Seger1987}. In other words, species do the same thing
every year, with only minimal response to environmental conditions. The
dry and variable environment of the sagebrush steppe has likely selected
for bet hedging species that can maintain function at low water
availability and have weak responses to high water availability. In so
doing, the dominant species in our ecosystem avoid ``boom and bust''
cycles, which corresponds to the weak effect of soil moisture on ANPP
(i.e., the Bayesian credible intervals for \(\beta\) in Eq. 1
overlapping zero).

Another strategy to deal with variable environmental conditions is
avoidance, which would also result in a consistent ecosystem functional
response between drought and control treatments. The perennial plants in
this cold desert ecosystem are tolerant to drought conditions
\citep[A.R. Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}.
For example, many of the perennial grasses in our focal ecosystem avoid
drought stress by growing early in the growing season \citep[A.R.
Kleinhesselink, personal observation]{Blaisdell1958}. Furthermore, the
dominant shrub in our focal ecosystem, \emph{Artemisia tripartita}, has
access to water deep in the soil profile thanks to a deep root system
\citep{Kulmatiski2017a}.

The weaker soil moisture-ANPP relationship we observed for the
irrigation treatment, evidenced by the less positive slope relative to
controls (Fig. 3A, right panel), has three non-exclusive explanations.
First, species' biomass production may have hit physiological thresholds
and/or become limited by other factors, such as nitrogen availability
\citep{LeBauer2008}.
\pba{PETER: DO WE KNOW ANYTHING ABOUT NITROGEN AT USSES? WANT TO ADD A SENTENCE HERE ABOUT HOW LIKELY THIS IS...}
Second, the less positive slope could be another indicator of bet
hedging in this community. Avoiding high biomass production in wet years
could benefit species in the long run if climate conditions are
variable. Third, the less positive slope may be a statistical artifact
due the variability of ANPP within years and our low sample size.

In conclusion, our results suggest the species in our focal plant
community are tolerant of drought conditions and bet hedgers in wet
conditions. Longer time series of chronic precipitation alteration may
reveal plant community shifts that we did not observe
\citep[e.g.,][]{Wilcox2016}, in which case species that do not bet hedge
may gain prominence and dominate the ecosystem functional response. Our
results suggest compositional shifts would have the largest impact at
high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.
